|

### 2021-12-16 05-Chp-CPRD-Analysis.Rmd
Range of updates

Electronic health record (EHR) databases are large collections of medical data, used to manage patient administration and care. Under these systems, whenever a patient attends their GP, their clinical data is recorded in a central database using a standardised coding system. These databases have several advantages over traditional methods of data collection, including the number of people they contain and the relatively low cost of data collection via routine care. This is particularly important when studying diseases such as dementia, which may begin to develop in patients long before symptoms are seen.

@@

This analysis makes use of the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of more than 3 million people from general practices across England. Using this data, the analysis presented in this chapter examined whether treatments which lower cholesterol levels (also known as lipid-regulating agents or LRA) of which statins are a prime example, affect the risk of all-cause dementia and related outcomes (Alzheimer's disease, vascular dementia and other dementias).

@@

Little evidence for an effect of lipid-regulating agents effect on the risk of Alzheimer's disease was found, with the exception of a slightly increased risk in those prescribed a certain type of lipid-regulating agent called fibrates. In contrast, I found an increased risk of vascular and other (i.e. non-Alzheimer's) dementia was associated with lipid-regulating agent use.

@@

This increased risk in outcomes with a vascular element (e.g. vascular dementia) is unexpected, and is very likely to be due to the presence of bias in the analysis. This bias, called "confounding by indication", is caused when those who are prescribed a statin are more at risk of vascular dementia for a range of reasons, which makes it appear as if statins are harmful. Despite this limitation, the analysis presented provides an important source of information which will be used in later chapters.

@@

In this chapter, I present an analysis of a large population-based electronic health record dataset to investigate the relationship between lipid-regulating agent (LRA) use and dementia outcomes. The analysis aims to address two important limitations of the current evidence base as identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading).

Firstly, it explicitly examines vascular dementia as an outcome. The systematic review presented in the previous chapter identified an evidence gap around the effect of lipid-regulating agents on the risk of vascular dementia. As triangulation exercises require as many diverse sources of evidence as possible, this analysis provides a source of information on this outcome.

Secondly, and in a similar vein, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia as identified by the systematic review. Specifically, this involved a concerted effort to address immortal time bias through use of a Cox proportional hazards analysis, incorporating a time-varying treatment indicator.[@suissa2008] By employing this approach, the analysis provides an evidence source at risk of a distinct bias. The results from this analysis will be incorporated into the triangulation exercise presented in Chapter \@ref(discussion-heading).

@@

<!-- QUESTION If protocol was written by Venexia, how do I reference this? -->

<!-- TODO Add amendments to Appendix -->

@@

Previously known as the General Practice Research Database, the Clinical Practice Research Datalink (CPRD) is a large population-based EHR database.[@herrett2015] The database has been collecting primary care data from participating practices across England since 1987.[@williams2012; @wood2001revitalizing] It contains the primary care records for more than 10 million primary care patients in England, and is broadly representative of the UK population in terms of age, sex and ethnicity.[@herrett2015; @mathur2014]

To avoid the ambiguity of interpreting free-text clinical notes and to allow for easy analysis of the resulting data, the CPRD primarily collects data using a predefined coding system known as Read codes.[@booth1994] All clinical events, including clinical test results and diagnoses, can be identified by a specific Read code. The codes use a nested approach (see Table \@ref(tab:readExample-table)), with the initial characters defining broad diagnostic topics (e.g. Eu... - Mental and behavioural disorders), while subsequent characters provide additional information on the specific condition diagnosed (e.g. Eu001 - Dementia in Alzheimer's disease with late onset).

@@

<!--------------------------------------------------------------------------->
(ref:readExample-caption) __Example of CPRD Read code hierarchy__ - Broad topics are specified using the initial two alpha-numeric characters of the Read code, while subsequent characters are used to define specific conditions and context. The example shown illustrates how "Dementia in Alzheimer’s disease with late onset" (_Eu001_) is nested under the top-level of "Mental disorders" (_Eu..._).

@@

The index events, exposures and outcomes used in this analysis were identified using predetermined code lists, which are available for inspection from the archived repository accompanying this analysis (data/code availability is discussed in Section \@ref(cprd-data-avail)).

<!-- TODO ensure code list position is accurately documented in Appendix -->

@@

This analysis included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data (as defined by the CPRD). Records pre-dating the 1995 cut-off were included in the original CPRD extract obtained for this analysis. However, these older records were excluded from the analysis as data quality and reliability are thought to be higher after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort 1 January 1996.

@@

Participants were ineligible for the cohort if they were less than 40 years of age (as these patients are less likely to be prescribed a LRA), had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with an outcome of interest before or on the date of the index event (i.e. had less than one full day of follow-up).

@@

A participants drug class was assigned based on their first recorded prescription, and any drug switching was ignored in an effort to mimic an intention-to-treat approach. We did however examine how often the initial drug class altered according to one of three criteria:

@@

<!-- TODO Need to deal with the fact that the ordering is not accounted for, which could be critised -->

<!----------------------------------------------------------------------------->
(ref:decisionTreeFig-cap) __Decision tree for assigning dementia subtypes__ - Based on the presence of specific Read codes in the patient's record, a decision tree was used to classify dementia subtypes ("Probable" or "Possible" Alheimer's disease, vascular dementia, and other dementia). Note that an outcome of "probable" or "Possible" Alzheimer's disease (AD) requires the absence of any vascular outcome codes.

@@

A range of additional variables were included in the analysis. These were intended to address the different distributions of potential confounding factors between those who were prescribed a lipid-regulating agent and those who were not. These are discussed in detail below and summarised in Table \@ref(tab:covariateDef-table).

@@

<!-- TODO May need to reword this table slightly as taken from Venexia walkers -->

<!----------------------------------------------------------------------------->
(ref:covariateDef-caption) __Definitions of covariates adjusted for in the analysis__ - The code lists used to define the majority of these covariates were originally created for use in a previously published analysis.[@walker2020] while others were built on or adapted from previous published work,[@khan2010; @taylor2016; @wright2017].

@@

Demographic covariates adjusted for included age and gender. Age in years was calculated at date of entry into the cohort, using the 1st of January of a patient's birth year (the exact date of birth was not provided by CPRD), and was adjusted for via its use as the time axis for the Cox model (see Section \@ref(cprd-time-axis)). Socioeconomic status was proxied using the Index of Multiple Deprivation (IMD) 2010. The IMD draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. To help preserve patient privacy, IMD score is only available from the CPRD in twentiles, with 1 indicating the least deprived and 20 indicating the most deprived. Smoking and alcohol use was determined at index, and was categorised as current, former, or never use.

Body mass index (a summary measure calculated as $weight/height^2$), baseline total cholesterol and baseline LDL cholesterol measures were obtained, using the last recorded value prior to the index date. A variable indicating grouped year of entry into the cohort (<=2000, 2001-2005, 2006-2010, >2010) was included to allow for changes in prescribing trends across the lifetime of the cohort. To assess healthcare utilisation, I adjusted for the average annual number of consultations between the beginning of a patient's data and their entry into the cohort.

Finally, presence of a range of related conditions at baseline were accounted for, including cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. In addition to adjusting for these covariates individually, a Charlson co-morbidity index (CCI) score was calculated for each participant. The CCI is a weighted index that uses presence and severity of a number of conditions to enable adjustment for the general health of a participant in terms of their mortality risk.[@charlson1987new] The conditions considered under the index are: AIDS; cancer; cerebrovascular disease; chronic pulmonary disease; congestive heart disease; dementia; diabetes; diabetes with complications; hemiplegia; metastatic tumour; mild liver disease; moderate liver disease; myocardial infarction; peptic ulcer disease; peripheral vascular disease; renal disease; and rheumatological disease. Inclusion of this index allowed me to attempt to adjust for the general health of patients included in the analysis.

code lists for all covariates can be found in the archived data repository accompanying this analysis (see Section \@ref(cprd-data-avail)).

@@

<!-- IDEA Add more detail here on the missing data approach -->

In this analysis, missing data was handled using a multiple imputation approach.[@sterne2009] Variables with missing observations were identified for inclusion in the imputation model. Nominal variables with missing values were modelled using multinomial logistic regression, while continuous variables were modelled using linear regression. As per best practice, all variables used in the analytic model, including the outcome, were included in the imputation model.[@moons2006] Using the MICE (Multiple Imputation by Chained Equations) command in STATA16, 20 imputed datasets were created.

Missing data was only considered problematic for variables where a numerical test result was expected (e.g. BMI), or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary, as absence of a code for other treatments or conditions (e.g. statin use or dementia) was assumed to indicate absence of the treatment/condition rather than being considered missing.[@wells2013strategies]

@@

* $t$ is the survival time;
* $h(t)$ is the hazard function; and
* $x_1,x_2,...,x_p$ are the covariates which determine the hazard function, while $b_1,b_2,...,b_p$ are the coefficients for each covariate.
* $h_o(t)$ is the baseline hazard - when all $x_i$ are zero, the $exp()$ function resolves to 1.

@@

<!-- IDEA Bit on why Cox models are slightly different from others, due to proportional hazards assumptions. Test for non-proportionality are therefore very important, as models which fail-->

<!-- IDEA Need to describe accounting for clustering here: id it something I did? See other thesis for clustering by practice -->

@@

Immortal time bias describes two distinct but related types of bias, considered here in relation to statin use. The first presentation, the selection bias aspect (Panel A, Figure \@ref(fig:immortalTimeBias)), occurs when time prior to statin initiation is excluded leading to the statin and control groups being followed up from different time points.[@levesque2010] Following the example displayed in the figure, the unexposed group are followed from the date of an index event (e.g. diagnosis of hypercholesterolemia), while the statin group is followed from date of a statin initiation. In this scenario, the time between the index event and statin initiation is missing, and any events that occur in the exposed group prior to the prescription will be inappropriately excluded from the analysis.

The second presentation of immortal time bias is as a type of misclassification bias (Panel B, Figure \@ref(fig:immortalTimeBias)). It occurs when the exposure time prior to statin initiation, and any events occurring within it, is inappropriately assigned to the exposed group.

<!-- TODO Reread this section and compare with the figure! -->

@@

This second presentation appears to be common in the existing literature on the relationship of statins and dementia, as several of the studies included in the systematic review performed as part of this thesis were identified as being at risk of immortal time bias following formal risk of bias assessment using the ROBINS-I tool (see Section \@ref(risk-of-bias-res)).

<!-- TODO May need to crosslink a bit better -->

Following a recommended approach to addressing the second form of immortal time bias, I employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010] Under this approach, all patients are followed from a common index date, defined as earliest of: (a) date of raised cholesterol test results; (b) hypercholesterolemia diagnosis; or (c) LRA prescription. Patients start in the unexposed group and contribute time-at-risk until they are prescribed a lipid-regulating agent, at which point they move into the exposed group. Note, patients for whom prescription of a lipid-regulating agent was the index event only contribute time to the exposed group (i.e. they enter the cohort and move into the exposed group on the same day).

@@

<!-- IDEA Finish paragraph from fragments below -->

<!-- In Cox models, there is the option to -->

<!-- Age vs time-since entry -->

<!-- Expand on use of age as the time scale, but describe the problems this introduced -->

As part of a Cox proportional hazard model, there is the option to use either absolute time in cohort or participants age as the time scale of interest.[@lamarca1998; @gail2009; @pencina2007] A model using age as the time axis inherently accounts, or adjusts, for participants age as a potential confounder of the exposure-outcome relationship. As such, the main analyses presented all used age as the time axis.

<!-- TODO Will need to expand on choice of time axis here -->

<!-- IDEA In order to explore whether the choice of time axis affects the results of the analysis,I ran two simple models, one using participants age as the time-scale and a second using absolute time-since-entry and adjusting for participant age. -->

@@

Negative outcomes are defined as those without a likely causal path between the exposure and outcome (see Figure \@ref(fig:negativeOutcome) for a directed acyclic graph, or DAG, describing an ideal negative outcome). Conversely, positive control outcomes are those with a known causal association with the exposure of interest, ideally sourced from large well conducted randomised controlled trials. Positive control outcomes are useful in observational epidemiology, as if the analysis can reproduce a known result for the control outcome, confidence in the result for the outcome of interest is increased. Due to the wealth of data available on statins as a lipid-regulating agent, I chose three control outcomes in reference to this drug class: back pain (negative control), ischaemic heart disease (positive protective control), and Type 2 diabetes (positive harmful control).

&nbsp;<!----------------------------------------------------------------------->  
(ref:negativeOutcome-cap) __Causal diagram for ideal negative outcome__ - The directed acyclic graph shows the relationship between the exposure $X$, outcome $Y$, confounders (measured $C$ and unmeasured $U$) and an ideal negative outcome $N$. Note the absence of any arrow between $X$ and $N$. In this scenario, any association observed between $X$ and $N$ is due to the presence of uncontrolled confounders $U$ (assuming $C$ has been adjusted for).

@@

Despite observational analyses suggesting a link between statins and muscular pain (as opposed to more serious complications such as myopathy),[@selva-ocallaghan2018] systematic reviews of the adverse events of statin use[@collins2016] and N-of-1 trials explicitly exploring the association of statin use with muscle pain[@herrett2021] have found little evidence supporting an effect. This suggests that muscular backpain would be suitable for use as a negative control outcome in this analysis. Under this approach, if statin use is found to be associated with muscular backpain in this analysis, this suggests the presence of residual confounding and reduces my confidence in the results for the dementia outcomes.

Similarly, incident ischemic heart disease and Type 2 diabetes were included as a protective and harmful positive control outcome, respectively. The protective effect of lipid-lowering treatment, via statins, on the risk of ischemic heart disease is well-established, [@collins2016] while there is growing evidence for an increased risk of Type 2 diabetes with statin use.[@collins2016; @macedo2014; @smit2020] Similar to the negative outcome, if the analysis strategy can reproduce these known associations, this will provide evidence that potential confounders have been sufficiently adjusted for.

<!-- QUESTION - should those with baseline cardiovascular conditions be excluded from the IHD/MI analysis? -->

@@

<!-- IDEA Make sure to include sex as a sensitivity analysis, given that that is the purpose of the IPD analysis -->

@@

<!-- IDEA Could frame this as providing evidence that the effect is not due to statins acting on lipid levels, as then would not expect difference (unless all very hard acting statins are in the lipophilic type) -->

As detailed in the introduction, the properties of statins may be important due to the ability of lipophilic statins to cross the blood brain barrier (see Section \@ref(intro-statins)).[@sierra2011] As such, I expected that any effects of statins on dementia outcomes would be stronger in the lipophilic as compared to the hydrophilic statin subgroup. To investigate this, I further stratified the statin exposure group into lipophilic (Atorvastatin, Lovastatin, Simvastatin, Cerivastatin) and hydrophilic (Pravastatin, Rosuvastatin, Fluvastatin) statins.

@@

As part of an exploratory analysis of the effect of the choice of code lists on the analysis, I created an alternative Alzheimer's disease and non-Alzheimer's dementia outcome using code lists from a published study by Smeeth _et al_.[@smeeth2009] The intended purpose of this analysis was to assess the robustness of my results to the choice of code list.

@@

<!-- QUESTION Describe impact of SSA table - is the fact that many people stopped statin treatment (~10%) an issue? Comment specifically in the context of a -->

The distribution of baseline characteristics across the remaining seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table). Note due to the experimental design, the median year of entry is expected to be later for those not prescribed an LRA, as this exposure group is more likely to include those who entered into the cohort towards the end of study window (as these people had less follow-up time in which to be prescribed an LRA).

@@

Full covariate information was available for `r missingtext`. Six key variables had some missing data: IMD 2010 score was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`; and baseline LDL cholesterol was missing for `r ldltext`.

@@

For each outcome, the overall "Any drug" estimate was driven by the statin subgroup, based on its large size relative to the other drug classes.

@@

<!-- TODO Check if reason for discrepnacy in terms of total numbers is that patients were excluded for having a first drug prior to on the index date -->

@@

The results of the series of sensitivity analyses performed are described in the following sections.

@@

In almost all cases, the use of imputed data resulted in a marginal attenuation of the effects observed when using a complete cases analysis. It should be noted that due to the large amount of missing data (e.g. `r ldltext` were missing a baseline LDL cholesterol measure), the number of participants included in the complete case analysis was substantially smaller than that included when using imputed data. In this case, though the overall position of the effect estimates does not change substantially when using the imputed dataset, there is a noticeable gain in power.[@sterne2009]

@@

<!----------------------------------------------------------------------------->
(ref:completeCaseFig-cap) __Complete case vs. imputed data analysis__ Comparison of analyses using the complete case versus imputed cohorts, indicating broad agreement between the two approaches. Note that the analyses using imputed data gave more precise estimates due to the low proportion of patients with complete covariate data.

(ref:completeCaseFig-scap) Complete case vs. imputed data analysis

@@

Following the primary analysis, the fully adjusted model was used to estimate the effect of treatment with a statin on the two control outcomes of back pain (negative) and ischemic heart disease (positive). The results of this analysis are presented in Figure \@ref(fig:controlOutcomeFig).

@@

<!----------------------------------------------------------------------------->
(ref:controlOutcomeFig-cap) Results of control outcome analysis.

(ref:controlOutcomeFig-scap) Results of control outcome analysis

@@

The results of three models adjusted for age only, age and sex, and the full covariates respectively, are presented in Figure \@ref(fig:unadjustedComparisonFig). These models were used to estimate the impact of adjustment for additional covariates. Note that obtaining a completely unadjusted model is not possible, as age was used in the Cox model as the time scale.

>>>>

Electronic health record (EHR) databases are large collections of medical data, used to manage patient administration and care. Under these systems, whenever a patient attends their GP, their clinical data is recorded in a central database using a standardised coding system. These databases have several advantages over traditional methods of data collection, including the number of people they contain and the relatively low cost of data collection. This is particularly important when studying diseases such as dementia, which may begin to develop in patients long before symptoms are seen.

@@

This analysis makes use of the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of more than 3 million people from general practices across England. Using this data, the analysis presented in this chapter examined whether lipid-regulating agents (treatments which change lipid levels) affect the risk of all-cause dementia and related outcomes (Alzheimer's disease, vascular dementia and other dementias).

@@

Little evidence for an effect of lipid-regulating agents on the risk of Alzheimer's disease was found, with the exception of a slightly increased risk in those prescribed a certain type of lipid-regulating agent called fibrates. In contrast, I found an increased risk of vascular and other (i.e. non-Alzheimer's) dementia was associated with lipid-regulating agent use.

@@

This increased risk in outcomes with a vascular element (e.g. vascular dementia) is unexpected, and is very likely to be due to the presence of bias in my analysis. This bias, called "confounding by indication", is caused when those who are prescribed a statin are more at risk of vascular dementia for a range of reasons, which makes it appear as if statins are harmful. Despite this limitation, the analysis provides an important source of information which will be used in later chapters.

@@

In this chapter, I present an analysis of a large population-based electronic health record (EHR) dataset to investigate the relationship between lipid-regulating agent (LRA) use and dementia outcomes. The analysis aims to address two important limitations of the current evidence base as identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading).

Firstly, it explicitly examines vascular dementia as an outcome. The systematic review presented in the previous chapter identified an evidence gap around the effect of lipid-regulating agents on the risk of vascular dementia. As triangulation exercises require as many diverse sources of evidence as possible, this analysis provides an additional source of information on this outcome.

Secondly, and in a similar vein, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia as identified by the systematic review. Specifically, this involved a concerted effort to address immortal time bias through use of a Cox proportional hazards analysis, incorporating a time-varying treatment indicator.[@suissa2008] By employing this approach, the analysis provides an evidence source at risk of a distinct bias, making it useful to the triangulation exercise presented in Chapter \@ref(tri-heading).

@@

@@

Previously known as the General Practice Research Database, the Clinical Practice Research Datalink (CPRD) is a large population-based EHR database.[@herrett2015] The database has been collecting primary care data from participating practices across England since 1987.[@williams2012; @wood2001revitalizing] It contains the records of more than 10 million primary care patients in England, and is broadly representative of the UK population in terms of age, sex and ethnicity.[@herrett2015; @mathur2014]

To avoid the ambiguity of interpreting free-text clinical notes and to allow for easy analysis of the resulting data, the CPRD primarily collects data using a predefined coding system known as Read codes.[@booth1994] All clinical events, including  test results and diagnoses, can be identified by a specific Read code. The codes use a nested approach (see Table \@ref(tab:readExample-table)), with the initial characters defining broad diagnostic topics (e.g. Eu... - Mental and behavioural disorders), while subsequent characters provide additional information on the specific condition diagnosed (e.g. Eu001 - Dementia in Alzheimer's disease with late onset).

@@

<!--------------------------------------------------------------------------->
(ref:readExample-caption) __Example of CPRD Read code hierarchy__ - Broad topics are specified using the initial two alpha-numeric characters of the Read code, while subsequent characters are used to define specific conditions and context. The example shown illustrates how "Dementia in Alzheimer’s disease with late onset" (_Eu001_) is nested under the top-level of "Mental and behavioural disorders" (_Eu..._).

@@

The index events, exposures and outcomes used in this analysis were identified using predetermined code lists, which are available for inspection from the repository accompanying this analysis (data/code availability is discussed in Section \@ref(cprd-data-avail)).

<!-- TODO Make sure it is cited properly -->

@@

This analysis included all patients registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data (as defined by the CPRD). Records pre-dating the 1995 cut-off were provided as part of the original CPRD extract obtained for this analysis. However, these older records were excluded from the analysis because data quality and reliability are thought to be higher after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort 1 January 1996.

@@

Participants were ineligible for the cohort if they were less than 40 years of age (because these patients are less likely to be prescribed a LRA), had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficultly of assigning these to a single exposure group), or were diagnosed with an outcome of interest before or on the date of the index event (i.e. had less than one full day of follow-up).

@@

A participants drug class was assigned based on their first recorded prescription, and any drug switching was ignored in an effort to mimic an intention-to-treat approach. I did however examine how often the initial drug class altered according to one of three criteria:

@@

<!----------------------------------------------------------------------------->
(ref:decisionTreeFig-cap) __Decision tree for assigning dementia subtypes__ - Based on the presence of specific Read codes in the patient's record, a decision tree was used to classify dementia subtypes ("Probable" or "Possible" Alzheimer's disease, vascular dementia, and other dementia). Note that an outcome of "probable" or "possible" Alzheimer's disease (AD) requires the absence of any vascular outcome codes.

@@

A range of additional variables were included in the analysis. These covariates were adjusted for in the analysis in an attempt to address the different distributions of potential confounding factors between those who were prescribed a lipid-regulating agent and those who were not. These are discussed in detail below and summarised in Table \@ref(tab:covariateDef-table).

@@

<!----------------------------------------------------------------------------->
(ref:covariateDef-caption) __Definitions of covariates adjusted for in the analysis__ - The code lists used to define the majority of these covariates were originally created for use in a previously published analysis.[@walker2020] while others were built on or adapted from previous published work.[@khan2010; @taylor2016; @wright2017]

@@

Demographic covariates adjusted for included age and gender. Age in years was calculated at date of entry into the cohort, using the 1st of January of a patient's birth year (the exact date of birth was not provided by CPRD), and was adjusted for via its use as the time axis for the Cox model (see Section \@ref(cprd-time-axis)). Socioeconomic status was proxied using the Index of Multiple Deprivation (IMD) 2010. The IMD draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. To help preserve patient privacy, IMD score is only available from the CPRD in twentiles, with 1 indicating the least deprived and 20 indicating the most deprived. Smoking and alcohol use was determined at index, and usage was categorised as current, former, or never.

Body mass index (a summary measure calculated as $\frac{weight}{height^2}$), baseline total cholesterol and baseline LDL cholesterol measures were obtained, using the last recorded value prior to the index date. A variable indicating grouped year of entry into the cohort (<=2000, 2001-2005, 2006-2010, >2010) was included to allow for changes in prescribing trends across the lifetime of the cohort. To assess healthcare utilisation, I adjusted for the average annual number of consultations between the beginning of a patient's data and their entry into the cohort.

Finally, presence of a range of related conditions at baseline were accounted for, including cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. In addition to adjusting for these covariates individually, a Charlson co-morbidity index (CCI) score was calculated for each participant. The CCI is a weighted index that uses presence and severity of a number of conditions to enable adjustment for the general health of a participant in terms of their mortality risk.[@charlson1987new] The conditions considered under the index are: AIDS; cancer; cerebrovascular disease; chronic pulmonary disease; congestive heart disease; dementia; diabetes; diabetes with complications; hemiplegia; metastatic tumour; mild liver disease; moderate liver disease; myocardial infarction; peptic ulcer disease; peripheral vascular disease; renal disease; and rheumatological disease. Inclusion of this index allowed me to adjust for the general health of patients included in the analysis.

Code lists for all covariates can be found in the archived data repository accompanying this analysis (see Section \@ref(cprd-data-avail)).

@@

In this analysis, missing data was handled using a multiple imputation approach.[@sterne2009] Variables with missing observations were identified for inclusion in the imputation model. Nominal variables with missing values were modelled using multinomial logistic regression, while continuous variables were modelled using linear regression. As per best practice, all variables used in the analytic model, including the outcome, were included in the imputation model.[@moons2006] Using the MICE (Multiple Imputation by Chained Equations) command in STATA 16,[@statacorp2019] 20 imputed datasets were created.

Missing data was only considered problematic for variables where a numerical test result was expected (e.g. BMI), or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary because absence of a code for other treatments or conditions (e.g. statin use or dementia) was assumed to indicate absence of the treatment/condition rather than being considered missing.[@wells2013strategies]

@@

* $t$ is the survival time;
* $h(t)$ is the hazard function;
* $x_1,x_2,...,x_p$ are the covariates which determine the hazard function, while $b_1,b_2,...,b_p$ are the coefficients for each covariate.; and
* $h_o(t)$ is the baseline hazard - when all $x_i$ are zero, the $exp()$ function resolves to 1.

@@

@@

Immortal time bias describes two distinct but related types of bias, considered here in relation to statin use (Figure \@ref(fig:immortalTimeBias)). The first presentation, the selection bias aspect (Panel A), occurs when time prior to statin initiation is excluded leading to the statin and control groups being followed up from different time points.[@levesque2010] The unexposed group are followed from the date of an index event (e.g. diagnosis of hypercholesterolemia), while the statin group is followed from date of a statin initiation. In this scenario, the time between the index event and statin initiation is missing, and any events that occur in the exposed group prior to the prescription will be inappropriately excluded from the analysis.

The second presentation of immortal time bias is as a type of misclassification bias (Panel B, Figure \@ref(fig:immortalTimeBias)). It occurs when the exposure time and events prior to statin initiation are inappropriately assigned to the statin group.

@@

This second presentation appears to be common in the existing literature on the relationship of statins and dementia. Several of the studies included in the systematic review performed as part of this thesis were identified as being at risk of immortal time bias following formal risk of bias assessment using the ROBINS-I tool (see Section \@ref(risk-of-bias-res)).

Following a recommended approach to address the second form of immortal time bias, I employed a time-varying treatment indicator to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010] Under this approach, all patients are followed from a common index date, defined as earliest of:

* date of raised cholesterol test results; 
* date of hypercholesterolemia diagnosis; or 
* date of LRA prescription.

Patients start in the unexposed group and contribute time-at-risk until they are prescribed a lipid-regulating agent, at which point they move into the exposed group. Note, patients for whom prescription of a lipid-regulating agent was the index event only contribute time to the exposed group (i.e. they enter the cohort and move into the exposed group on the same day).

@@

As part of a Cox proportional hazard model, there is the option to use either absolute time in cohort or participants' age as the time scale of interest.[@lamarca1998; @gail2009; @pencina2007] A model using age as the time axis inherently accounts, or adjusts, for participants age as a potential confounder of the exposure-outcome relationship. As such, the analyses presented all used age as the time axis.

@@

Negative outcomes are defined as those without a likely causal path between the exposure and outcome (see Figure \@ref(fig:negativeOutcome) for a directed acyclic graph, or DAG, describing an ideal negative outcome). Conversely, positive control outcomes are those with a known causal association with the exposure of interest, ideally sourced from large well conducted randomised controlled trials. Positive control outcomes are also useful in observational epidemiology because if the analysis can reproduce a known result for the control outcome, confidence in the result for the outcome of interest is increased. Due to the wealth of data available on statins as a lipid-regulating agent, I chose three control outcomes in reference to this drug class: back pain (negative control), ischemic heart disease (positive protective control), and Type 2 diabetes (positive harmful control).

&nbsp;<!----------------------------------------------------------------------->  
(ref:negativeOutcome-cap) __Causal diagram for an ideal negative outcome__ - The directed acyclic graph shows the relationship between the exposure $X$, outcome $Y$, confounders (measured $C$ and unmeasured $U$) and an ideal negative outcome $N$. Note the absence of any arrow between $X$ and $N$. In this scenario, any association observed between $X$ and $N$ is due to the presence of uncontrolled confounders $U$ (assuming $C$ has been adjusted for).

@@

Despite observational analyses suggesting a link between statins and muscular pain (as opposed to more serious complications such as myopathy),[@selva-ocallaghan2018] an effect was not observed in either systematic reviews of adverse events of statin use[@collins2016] or N-of-1 trials explicitly exploring the association of statin use with muscle pain.[@herrett2021] Based on this evidence, I used muscular backpain as a negative control outcome in this analysis. Under this approach, if statin use is found to be associated with muscular backpain in this analysis, this suggests the presence of residual confounding and reduces my confidence in the results for the dementia outcomes.

Similarly, incident ischemic heart disease and Type 2 diabetes were included as protective and harmful positive control outcome, respectively. The protective effect of lipid-lowering via statins on the risk of ischemic heart disease is well-established, [@collins2016] while there is growing evidence for an increased risk of Type 2 diabetes with statin use.[@collins2016; @macedo2014; @smit2020] Similar to the negative outcome, if the analysis strategy can reproduce these known associations, this will provide evidence that potential confounders have been sufficiently adjusted for.

@@

@@

As detailed in the introduction, the properties of statins may be important due to the ability of lipophilic statins to cross the blood brain barrier (see Section \@ref(intro-statins)).[@sierra2011] As such, I expected that any effects of statins on dementia outcomes would be stronger in the lipophilic versus the hydrophilic statin subgroup. To investigate this, I further stratified the statin exposure group into lipophilic (Atorvastatin, Lovastatin, Simvastatin, Cerivastatin) and hydrophilic (Pravastatin, Rosuvastatin, Fluvastatin) statins.

@@

To explore the impact of the code lists used to define dementia outcomes, I created alternative Alzheimer's disease and non-Alzheimer's dementia outcomes using code lists from a published study by Smeeth _et al_.[@smeeth2009] The intended purpose of this analysis was to assess the robustness of my results to the choice of code list.

@@

The distribution of baseline characteristics across the remaining seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table). Note due to the experimental design, the median year of entry is expected to be later for those not prescribed an LRA. This is because the unexposed group is more likely to include those who entered into the cohort towards the end of study window, and so had less follow-up time in which to be prescribed an LRA.

@@

Full covariate information was available for `r missingtext`. Six key variables had some missing data: IMD 2010 score was missing for `r imdtext` because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`; and baseline LDL cholesterol was missing for `r ldltext`.

@@

For each outcome, the overall "Any drug class" estimate was driven by the statin subgroup, based on its large size relative to the other drug classes.

@@

@@

The results of the sensitivity analyses are described in the following sections.

@@

In almost all cases, the use of imputed data resulted in a marginal attenuation of the effects observed when using a complete case analysis. It should be noted that due to the large amount of missing data (e.g. `r ldltext` were missing a baseline LDL cholesterol measure), the number of participants included in the complete case analysis was substantially smaller than that included when using imputed data. In this case, though the overall position of the effect estimates does not change substantially when using the imputed dataset, there is a noticeable gain in power.[@sterne2009]

@@

<!----------------------------------------------------------------------------->
(ref:completeCaseFig-cap) __Sensitivity analysis: complete case vs. imputed data__ - Comparison of analyses using the complete case versus imputed cohorts, indicating broad agreement between the two approaches. Note that the analyses using imputed data gave more precise estimates due to the low proportion of patients with complete covariate data.

(ref:completeCaseFig-scap) Sensitivity analysis: complete case vs. imputed data

@@

The fully adjusted model was also used to estimate the effect of treatment with a statin on three control outcomes: back pain (negative), ischemic heart disease (positive protective) and Type 2 diabetes (positive harmful). The results of this analysis are presented in Figure \@ref(fig:controlOutcomeFig).

<!----------------------------------------------------------------------------->
(ref:controlOutcomeFig-cap) __Sensitivity analysis: control outcomes__ Effect estimates obtained from the fully-adjusted time-varying model with participant age as the time scale for three control outcomes: backpain (negative), ischemic heart disease (positive protective), and diabetes (positive harmful).

(ref:controlOutcomeFig-scap) Sensitivity analysis: control outcomes

@@

The results of three models adjusted for age only, age and sex, and full covariates respectively, are presented in Figure \@ref(fig:unadjustedComparisonFig). These models were used to estimate the impact of adjustment for additional covariates. Note that obtaining a completely unadjusted model is not possible because age was used in the Cox model as the time scale.

Adjustment for additional covariates beyond age and sex had a limited impact on the observed effect estimates, with the exception of the probable AD outcome. In this case, adjustment for the full set of covariates attenuated the protective effect observed when adjusting only for age and sex to the null.

@@

<!----------------------------------------------------------------------------->
(ref:unadjustedComparisonFig-cap) __Sensitivity analysis: covariate comparison__ - The results obtained using three different sets of covariates (age only, age + sex, all covariates) are shown for each dementia outcome. Note that the x-axis cutoffs (0.5,2) are different compared to other plots (0.3,3) to enable greater comparison between the different models.

(ref:unadjustedComparisonFig-scap) Sensitivity analysis: covariate comparison
